---
title: "Preprocessing"
author: "Mbouamboua Yvon<BR>mbouamboua@ipmc.cnrs.fr"
output:
  html_notebook:
    fig_caption: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
date: 'Last update: `r format(Sys.time())`'
---

# Defining working directory and subfolders

```{r setup}

dir.main = "/data/data_mbouamboua/projects/human_olfactory_cf"
dir.data = file.path(dir.main, "data")
h5.mtx <- "filtered_feature_bc_matrix.h5"
file.name <- list("G3_M2_M9","G1_M7_M10","M1_M5_M6","D571_572_573","D574_575_576_577")
dir.results <- file.path(dir.main, paste0("000-preprocessing"))
result.folders <- list("rdata","seurat_data", "tables", "plots")

dir.path <- c()
for (folder in result.folders) {
  dir.path[folder] <- file.path(dir.results, folder)
  if(!dir.exists(dir.path[folder])){
   dir.create(dir.path[folder], showWarnings = FALSE, recursive = TRUE) 
  }
}

```

# Load R packages and helper functions

```{r package_function}

suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratDisk)
  library(ggplot2)
  library(dplyr)
  library(pheatmap)
  library(ComplexHeatmap)
  library(ComplexUpset)
  library(stringr)
  library(patchwork)
})

source(file.path("/data/data_mbouamboua/utils/00-import.R"))

```



## Create seurat object


```{r}

for (i in 1:length(file.name)) {
  message("Create", file.name[[i]], "seurat object")
  scmat <- Read10X_h5(file.path(dir.data, file.name[[i]], h5.mtx))
  sobj <- CreateSeuratObject(counts = scmat, project = file.name[[i]])
  message("Save", file.name[[i]], "seurat object")
  saveRDS(sobj, file.path(dir.results, "/seurat_data/", filename = paste0(file.name[[i]], ".rds")))
}

```


## Demultiplexing: add demuxafy souporcell+vireo+scds combined results

```{r}

seurat.list <- list()

for (i in 1:length(file.name)) {  
  
  file_path = list.files(path = file.path(dir.results, "/seurat_data/"), pattern =  paste0(file.name[[i]], ".rds"), full.names = T, recursive = F)
  sobj <- readRDS(file_path)
  dmx <- read.delim(file.path(dir.main, "WS_demuxafy/", file.name[[i]],"combine_majoritySinglet/combined_results_w_combined_assignments.tsv"))
  dmx$barcode_seurat = dmx$Barcode
  rownames(dmx) <- dmx$barcode_seurat
  seurat.list[[file.name[[i]]]] <- AddMetaData(sobj, dmx)
  seurat.list[[file.name[[i]]]][['donor']] <- paste0(seurat.list[[file.name[[i]]]]@meta.data$MajoritySinglet_Individual_Assignment,"_", seurat.list[[file.name[[i]]]]@meta.data[,"orig.ident"])  
  rm(sobj)
}

unique(seurat.list[[1]]@meta.data)

```




## Add metadata

```{r}

seurat.list[["G1_M7_M10"]][["kaftrio"]]=""
seurat.list[["G1_M7_M10"]]@meta.data[seurat.list[["G1_M7_M10"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor0","donor1","donor2"),"kaftrio"] <-"Yes"
seurat.list[["G1_M7_M10"]]@meta.data[seurat.list[["G1_M7_M10"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("doublet"),"kaftrio"] <-"doublet"
seurat.list[["G1_M7_M10"]]@meta.data[seurat.list[["G1_M7_M10"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("unassigned"),"kaftrio"] <-"unassigned"

seurat.list[["G1_M7_M10"]][["sample"]]=""
seurat.list[["G1_M7_M10"]]@meta.data[seurat.list[["G1_M7_M10"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor0"),"sample"] <-"M10"
seurat.list[["G1_M7_M10"]]@meta.data[seurat.list[["G1_M7_M10"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor1"),"sample"] <-"M7"
seurat.list[["G1_M7_M10"]]@meta.data[seurat.list[["G1_M7_M10"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor2"),"sample"] <-"G1"
seurat.list[["G1_M7_M10"]]@meta.data[seurat.list[["G1_M7_M10"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("doublet"),"sample"] <-"doublet"
seurat.list[["G1_M7_M10"]]@meta.data[seurat.list[["G1_M7_M10"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("unassigned"),"sample"] <-"unassigned"

seurat.list[["G3_M2_M9"]][["kaftrio"]]=""
seurat.list[["G3_M2_M9"]]@meta.data[seurat.list[["G3_M2_M9"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor0"),"kaftrio"] <-"No"
seurat.list[["G3_M2_M9"]]@meta.data[seurat.list[["G3_M2_M9"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor1","donor2"),"kaftrio"] <-"Yes"
seurat.list[["G3_M2_M9"]]@meta.data[seurat.list[["G3_M2_M9"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("doublet"),"kaftrio"] <-"doublet"
seurat.list[["G3_M2_M9"]]@meta.data[seurat.list[["G3_M2_M9"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("unassigned"),"kaftrio"] <-"unassigned"

seurat.list[["G3_M2_M9"]][["sample"]]=""
seurat.list[["G3_M2_M9"]]@meta.data[seurat.list[["G3_M2_M9"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor0"),"sample"] <-"M9"
seurat.list[["G3_M2_M9"]]@meta.data[seurat.list[["G3_M2_M9"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor1"),"sample"] <-"G3"
seurat.list[["G3_M2_M9"]]@meta.data[seurat.list[["G3_M2_M9"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor2"),"sample"] <-"M2"
seurat.list[["G3_M2_M9"]]@meta.data[seurat.list[["G3_M2_M9"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("doublet"),"sample"] <-"doublet"
seurat.list[["G3_M2_M9"]]@meta.data[seurat.list[["G3_M2_M9"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("unassigned"),"sample"] <-"unassigned"

seurat.list[["M1_M5_M6"]][["kaftrio"]]=""
seurat.list[["M1_M5_M6"]]@meta.data[seurat.list[["M1_M5_M6"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor0","donor2"),"kaftrio"] <-"Yes"
seurat.list[["M1_M5_M6"]]@meta.data[seurat.list[["M1_M5_M6"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor1"),"kaftrio"] <-"No"
seurat.list[["M1_M5_M6"]]@meta.data[seurat.list[["M1_M5_M6"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("doublet"),"kaftrio"] <-"doublet"
seurat.list[["M1_M5_M6"]]@meta.data[seurat.list[["M1_M5_M6"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("unassigned"),"kaftrio"] <-"unassigned"

seurat.list[["M1_M5_M6"]][["sample"]]=""
seurat.list[["M1_M5_M6"]]@meta.data[seurat.list[["M1_M5_M6"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor0"),"sample"] <-"D0.M1.M5"
seurat.list[["M1_M5_M6"]]@meta.data[seurat.list[["M1_M5_M6"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor1"),"sample"] <-"M6"
seurat.list[["M1_M5_M6"]]@meta.data[seurat.list[["M1_M5_M6"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor2"),"sample"] <-"D2.M1.M5"
seurat.list[["M1_M5_M6"]]@meta.data[seurat.list[["M1_M5_M6"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("doublet"),"sample"] <-"doublet"
seurat.list[["M1_M5_M6"]]@meta.data[seurat.list[["M1_M5_M6"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("unassigned"),"sample"] <-"unassigned"

seurat.list[["D571_572_573"]][["kaftrio"]]=""
seurat.list[["D571_572_573"]]@meta.data[seurat.list[["D571_572_573"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor0","donor1","donor2"),"kaftrio"] <-"No"
seurat.list[["D571_572_573"]]@meta.data[seurat.list[["D571_572_573"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("doublet"),"kaftrio"] <-"doublet"
seurat.list[["D571_572_573"]]@meta.data[seurat.list[["D571_572_573"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("unassigned"),"kaftrio"] <-"unassigned"

seurat.list[["D571_572_573"]][["sample"]]=""
seurat.list[["D571_572_573"]]@meta.data[seurat.list[["D571_572_573"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor0"),"sample"] <-"D571"
seurat.list[["D571_572_573"]]@meta.data[seurat.list[["D571_572_573"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor1"),"sample"] <-"D572"
seurat.list[["D571_572_573"]]@meta.data[seurat.list[["D571_572_573"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor2"),"sample"] <-"D573"
seurat.list[["D571_572_573"]]@meta.data[seurat.list[["D571_572_573"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("doublet"),"sample"] <-"doublet"
seurat.list[["D571_572_573"]]@meta.data[seurat.list[["D571_572_573"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("unassigned"),"sample"] <-"unassigned"

seurat.list[["D574_575_576_577"]][["kaftrio"]]=""
seurat.list[["D574_575_576_577"]]@meta.data[seurat.list[["D574_575_576_577"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor0","donor1","donor2","donor3"),"kaftrio"] <-"No"
seurat.list[["D574_575_576_577"]]@meta.data[seurat.list[["D574_575_576_577"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("doublet"),"kaftrio"] <-"doublet"
seurat.list[["D574_575_576_577"]]@meta.data[seurat.list[["D574_575_576_577"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("unassigned"),"kaftrio"] <-"unassigned"

seurat.list[["D574_575_576_577"]][["sample"]]=""
seurat.list[["D574_575_576_577"]]@meta.data[seurat.list[["D574_575_576_577"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor0"),"sample"] <-"D574"
seurat.list[["D574_575_576_577"]]@meta.data[seurat.list[["D574_575_576_577"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor1"),"sample"] <-"D575"
seurat.list[["D574_575_576_577"]]@meta.data[seurat.list[["D574_575_576_577"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor2"),"sample"] <-"D576"
seurat.list[["D574_575_576_577"]]@meta.data[seurat.list[["D574_575_576_577"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("donor3"),"sample"] <-"D577"
seurat.list[["D574_575_576_577"]]@meta.data[seurat.list[["D574_575_576_577"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("doublet"),"sample"] <-"doublet"
seurat.list[["D574_575_576_577"]]@meta.data[seurat.list[["D574_575_576_577"]]@meta.data$MajoritySinglet_Individual_Assignment %in% c("unassigned"),"sample"] <-"unassigned"

```



# Plot demuxafy summary

```{r fig.width=6, fig.height=6}

plot_list <- list()

for (i in 1:length(file.name)) {
  
Idents(seurat.list[[file.name[[i]]]]) <- "sample"
stats <- Cluster_Stats(seurat.list[[file.name[[i]]]], group.by.var = "orig.ident", remove_total = T)

plot_list[[i]] <- ggplot(data = stats, mapping = aes(x = reorder(Cluster, -Number, decreasing = T), y = Number, fill = Cluster)) + 
  geom_bar(stat = "identity", fill = "#5694C7") + 
  theme_bw() +
  geom_text(aes(label = Number), vjust = -0.5, size=3) +
  ylim(0, max(stats$Number*1.1))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, colour = "black",size = 10),
        axis.text.y = element_text(colour = "black", size = 10),
        axis.title = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, vjust = 2, size = 10, face = "bold"),
        legend.position = "none") + 
  labs(title = paste0(file.name[[i]]),
       x = "",
       y = "Number of cells")
}

combined_plot <- patchwork::wrap_plots(plotlist = plot_list, guide = "collect", axes = "collect_y", axis_titles = "collect", ncol = 3)
print(combined_plot)
save_plot(plot = combined_plot, filename = file.path(dir.results, "plots/demuxafy"), width = 6, height = 6, formats = c("pdf","png"))

```


# Remove unassigned and doublet cells

```{r}

for (i in 1:length(seurat.list)) {
  
  Idents(seurat.list[[i]]) <- "MajoritySinglet_Individual_Assignment"
  seurat.list[[i]]@meta.data$MajoritySinglet_Individual_Assignment <- 
    seurat.list[[i]]@meta.data$MajoritySinglet_Individual_Assignment %>% tidyr::replace_na('doublet')
  unique(seurat.list[[i]]$MajoritySinglet_Individual_Assignment)
  idents <- unique(seurat.list[[i]]$MajoritySinglet_Individual_Assignment)
  idents <- setdiff(idents, c("unassigned", "doublet"))
  seurat.list[[i]] <- subset(seurat.list[[i]], idents = idents, invert = FALSE)
  print(unique(seurat.list[[i]]$MajoritySinglet_Individual_Assignment))
  
}

```


# Plot QC metrics of unfiltered seurat data

```{r fig.width=5, fig.height=4}

merged <- merge(x = seurat.list[[1]], y = seurat.list[2:length(seurat.list)], merge.data = TRUE)
merged <- PercentageFeatureSet(object = merged, pattern = "^MT-", col.name = "percent_mito")
merged@meta.data

Idents(merged) <- "sample"
p <- VlnPlot(object = merged, features = c("nCount_RNA","nFeature_RNA","percent_mito")) + NoLegend()
print(p)
save_plot(plot = p, filename = file.path(dir.results, "plots/preQC_metrics"), width = 5, height = 2, formats = c("pdf","png"))

```


# Filter low quality cells

```{r}

object_list <- list()
object_list <- SplitObject(object = merged, split.by = "sample")

results <- Run_Quality_Control(seurat_input = object_list, 
                               min_features = 200, 
                               max_mito = 5, 
                               filter_method = "none", 
                               filter_max_counts = F,
                               filter_doublets = F,
                               dir_results = file.path(dir.results))

```

# Plot QC metrics of filtered seurat data

```{r fig.width=5, fig.height=2}

seurat.list <- list()
seurat.list <- results$seurat_list
merged <- merge(x = seurat.list[[1]], y = seurat.list[2:length(seurat.list)], merge.data = TRUE)
rm(seurat.list)
Idents(merged) <- "sample"
p <- VlnPlot(object = merged, features = c("nCount_RNA","nFeature_RNA","percent_mito")) + NoLegend()
print(p)
save_plot(plot = p, filename = file.path(dir.results, "plots/postQC_metrics"), width = 5, height = 2, formats = c("pdf","png"))

```


# Seurat rPCA integration

```{r fig.width=8, fig.height=6}

seurat.list <- list()
seurat.list <- results$seurat_list
merged <- merge(x = seurat.list[[1]], y = seurat.list[2:length(seurat.list)], merge.data = TRUE)
seurat.list <- list()
seurat.list <- SplitObject(object = merged, split.by = "orig.ident")
rm(merged)

invisible(gc())
seurat.list <- lapply(X = seurat.list, FUN = function(x) {
    x <- NormalizeData(x)
    x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 5000)
})

invisible(gc())
features <- SelectIntegrationFeatures(object.list = seurat.list)
seurat.list <- lapply(X = seurat.list, FUN = function(x) {
    x <- ScaleData(x, features = features, verbose = FALSE)
    x <- RunPCA(x, features = features, verbose = FALSE)
})


invisible(gc())
anchors <- FindIntegrationAnchors(object.list = seurat.list, anchor.features = features, reduction = "rpca")
rm(seurat.list)
obj <- IntegrateData(anchorset = anchors)
rm(anchors)

obj <- FindVariableFeatures(obj, assay = "RNA", nfeatures = 5000)
features <- VariableFeatures(obj, assay = "RNA")
DefaultAssay(obj) <- "integrated"
obj <- ScaleData(obj)
obj <- RunPCA(obj, npcs = 30, features = features)
obj <- RunUMAP(obj, reduction = "pca", dims = 1:30, min.dist = 1.5)
obj <- FindNeighbors(obj, reduction = "pca", dims = 1:30)
obj <- FindClusters(obj, resolution = c(0.5))

invisible(gc())
saveRDS(obj, file.path(dir.results, "rdata/integrated.rds"))

DefaultAssay(obj) <- "RNA"
Dim_Plot(object = obj, group.by = "seurat_clusters", label.show = T, legend = T, theme.bw = F,figure.plot = T, pt.size = 0.5)

```


# Check cell clusters quality

```{r}

VlnPlot(object = obj, features = c("nCount_RNA","nFeature_RNA","percent_mito")) + NoLegend()

```



# Remove low quality cell cluster
The cluster 4 is low quality cells.

```{r fig.width=10, fig.height= 7}

Idents(obj) <- "seurat_clusters"
obj <- subset(obj, idents = c(4), invert = TRUE)

DefaultAssay(obj) <- "integrated"
obj <- ScaleData(obj)
obj <- RunPCA(obj, npcs = 30)
obj <- RunUMAP(obj, reduction = "pca", dims = 1:15, min.dist = 2.5)
obj <- FindNeighbors(obj, reduction = "pca", dims = 1:20)
obj <- FindClusters(obj, resolution = c(1.5))
Dim_Plot(obj, group.by = "seurat_clusters", label.show = T, pt.size = 1)
saveRDS(obj, file.path(dir.results, "rdata/integrated.rds"))

```


```{r session_info}

utils::capture.output(devtools::session_info())

```


